Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-08 paper.tex
Refactored sec 2 + new tikzed diagrams

\usepackage{xcolor}

\definecolor{lightgrey}{HTML}{FAFAFA}

@@

<?php
function make_titles($arg, $n) {
  $is = shuffle(array(1,2,3));
  while ($n < $is[0]) {
    echo "<h1>" . $arg . "</h1><br>";
    $n = $n + 1;
  }
}

@@

$name = "Hello " . $_POST['name'] .  "!";
if (!isset($name)) {
  echo "No name entered!";
} else {
  echo $name;
}
?>
\end{lstlisting}
		\caption{Snippet of HTML and PHP server-side web application code. \label{fig:example_code}}
    \end{subfigure}

@@

// #repeat n < shuffle(is)
  <h1>Headline</h1>
// #endrepeat

@@

\end{lstlisting}
		\caption{Approximation client paege output of the server-side web application code in Figure \ref{fig:example_code}. Variability as well as repetition annotations are represented by preprocessor directives.\label{fig:output_example}}
    \end{subfigure}
    \vspace{2mm}
    \caption{Illustration of static ouptut approximation using symbolic
    execution}
\end{figure}

% Workflow Graphic
%\begin{figure*}[t!]
%	\centerline{
%	\includegraphics[width=0.95\textwidth]{images-paper/workflow}
%	}
%	\vspace{2mm}
%	\caption{
%		Overview of the workflow for tools based on static output approximation for
%		% dynamic web applications.
%	}
%	\label{fig:approach}
%\end{figure*}

Figure \ref{fig:} shows two code listings. The snippet of server-side web application code in Figure \ref{} contains both HTML and PHP. First, the function make\_titles defined in lines XX – XY prints the string arg passed as an argument between zero and an arbitrary number of times. Second, the function is indirectly called using the built-in function call\_user\_func in line XX with the string “Headline” and 4 as arguments.  Finally, the input from the form in lines 1 – 4 is used: The variable greeting consists of three concatenated strings, “Hello ” the value of input field “name” and an exclamation mark. For this server-side code snippet there are two possible client page output variants: If the input field “name” is empty at runtime, line XX is executed and “No name entered!” is printed, or, if a name was entered, line XY is executed and the greeting constructed in line XZ is printed.

The following symbolic execution semantics for PHP features  based on [BCG] illustrate the necessary modifications made to handle the extent of ambiguity that comes with handling symbolic values. In addition to normal output, the output approximation contains preprocessor-like directives to express repeatable output as well as output variants. All approximated client page output corresponding to Figure \ref{1a} consisting of two different variants is illustrated in Figure \ref{1b}.

First, loops in the previous execution semantics are only executed once instead of an arbitrary number of times. Since a loop condition can contain symbolic values, it might not be feasible to determine when a loop terminates. Hence, any output constructed during the single loop iteration is highlighted (repeat … endrepeat) to possibly be repeated an arbitrary number of times. The output in line XX is marked to be repeated an arbitrary number of times, as the number of loop iterations cannot be determined statically. Second, a similar procedure applies to recursive functions. Once called, a recursive function is  also only executed to recursion depth one per call, and any subsequent invocation of that function returns a symbolic value. This, again, is due to the difficulty to determine the exact recursion depth similar to the number of loop iterations. Third, since the symbolic execution engine applies a static analysis, any form of user interaction providing input data as well as any form of input referred to from the deployment context of the 
Symbolic execution is an analysis technique not dedicated to a particular programming language. For dynamic web applications, in our case for web applications written in PHP, it can be used to approximate all possible client page output. Before we present in detail existing tools that aim to ease development and maintenance of PHP web applications, we illustrate how output approximation using symbolic execution is used by starting with an example.

B. Explanation of Figure 1
Figure \ref{fig:} shows two code listings. The snippet of server-side web application code in Figure \ref{} contains both HTML and PHP. First, the function make\_titles defined in lines XX – XY prints the string \$arg passed as an argument between zero and an arbitrary number of times. Second, the function is indirectly called using the built-in function call\_user\_func in line XX with the string “Headline” and 4 as arguments.  Finally, the input from the form in lines 1 – 4 is used: The variable \$greeting consists of three concatenated strings, “Hello ” the value of input field “name” and an exclamation mark. For this server-side code snippet there are two possible client page output variants: If the input field “name” is empty at runtime, line XX is executed and “No name entered!” is printed, or, if a name was entered, line XY is executed and the greeting constructed in line XZ is printed.

C. Introduction to symbolic execution of PHP code
The following symbolic execution semantics for PHP features  based on [BCG] illustrate the necessary modifications made to handle the extent of ambiguity that comes with handling symbolic values. In addition to normal output, the output approximation contains preprocessor-like directives to express repeatable output as well as output variants. All approximated client page output corresponding to Figure \ref{1a} consisting of two different variants is illustrated in Figure \ref{1b}. 
First, loops in the previous execution semantics are only executed once instead of an arbitrary number of times. Since a loop condition can contain symbolic values, it might not be feasible to determine when a loop terminates. Hence, any output constructed during the single loop iteration is highlighted (repeat … endrepeat) to possibly be repeated an arbitrary number of times. The output in line XX is marked to be repeated an arbitrary number of times, as the number of loop iterations cannot be determined statically. Second, a similar procedure applies to recursive functions. Once called, a recursive function is  also only executed to recursion depth one per call, and any subsequent invocation of that function returns a symbolic value. This, again, is due to the difficulty to determine the exact recursion depth similar to the number of loop iterations. Third, since the symbolic execution engine applies a static analysis, any form of user interaction providing input data as well as any form of input referred to from the deployment context of the web application, i.e. configuration files or databases, is substituted by a symbolic value. The context variable \$\_POST represents input submitted to the server and, hence, is symbolic as every program input is per definition substituted by symbolic values. Finally, the output in lines and XX and XY respectively are marked as alternatives (if … else … endif), as symbolic execution of lines XX – YY in Figure 1a will explore both different branches.

@@

\textcolor{blue}{Bener}

@@

In the following we describe in detail our implementation procedure, practical problems we encountered and whether these are limiting scalability conceptually or just require engineering effort to be tamed.

\subsection{Towards Scalable Output Approximation?}
We re-implement\footnote{\url{https://github.com/smba/oak}} the existing symbolic execution semantics of \cite{Nguyen:2014:BCG:2635868.2635928} with the goal to support modern large-scale dynamic web applications, or at least to better understand possible limitations. For reference, we choose to aim to support  \textsf{WordPress}\footnote{\url{https://github.com/WordPress/WordPress}}, a popular Content Management System of reasonable size that is still maintained. Moreover, our implementation uses parts of Quercus\footnote{\url{http://quercus.caucho.com/}}, a pure Java implementation of the PHP interpreter, which is tested against the \textsf{WordPress} test suite. We choose a test-driven approach for our implementation with regression tests and continuous integration. Our test cases are based on both custom examples for different language features as well as \textsf{SchoolMate}, a small-scale system, and, at least in parts, \textsf{WordPress}, whose output we manually validate and then automatically re-check for each test run.

@@

\subsection{Experiment Setup} 
For the evaluation we symbolically executed a selection of PHP systems. We selected a corpus of twelve PHP systems with regard to system size and recency as the corpus of the case study for \cite{Nguyen:2014:BCG:2635868.2635928} only contained small-scale systems that are not maintained any more. The full list of PHP systems is shown in Table \ref{corpus}. Our selection of PHP systems includes
%\begin{enumerate}[a)]
\begin{compactitem}
\item four small-scale systems that selected from the previously mentioned case study in order to compare our results with the previous symbolic execution engine \cite{Nguyen:2014:BCG:2635868.2635928},
\item three recent large-scale systems selected from the case study corpus of \cite{Hills:2013:ESP:2483760.2483786}, and
\item four small-scale systems selected from a list of recent Content Management Systems \cite{codegeekz}.
%umerate}

\end{compactitem}
As for for the last two cases, we limited our selection since the parser we
used did not support all language features used. For the experimental evaluation we symbolically executed per system each script file or entry point (files with a \texttt{.php}, \texttt{.inc}, \texttt{.bit} or \texttt{.module} extension). All following measurements are cumulated per system for each entry point.

\begin{table*}[t]
\centering 
	\begin{tabular}{lrp{4cm}rlrr}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{Classification} & \textbf{SLOC} & \textbf{\#files} & \textbf{\#OCs} & \textbf{\#includes}
	\\
	\midrule
	AddressBook & 8.2.5.2 & 239 & 51,907 & Adress Administration  & XXX & XXX\\
	SchoolMate & 1.5.4 & 65 & 8,118 & School Administration  & XXX & XXX\\
	TimeClock & 1.04 & 63 & 20,800 & XXX  & XXX & XXX\\
	WebChess & 1.0.0 & 28 & 5,219 & XXX  & XXX & XXX\\
	\midrule
	Drupal & 7.5.0 & 125 & 52,464 & Content Management System  & XXX & XXX\\
	phpBB & 3.1.9 & 1,398 & 327,371 & Bulletin Board  & XXX & XXX\\
	phpMyAdmin & 4.6.3 & 871 & 303,582 & Database Administration  & XXX & XXX\\
	\midrule
	Anchor & 0.12.1 & 201 & 15,054 & Content Management System  & XXX & XXX\\
	Kirby & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	Automad & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	Monstra & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	Nibbleblog & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

The coverage results are illustrated in Figure \ref{coverage}. We could replicate high coverage results for the four small-scale systems with a reach coverage and output coverage over 80 percent respectively. For the three modern/large-scale systems we measured poor coverage with reach coverage ranging from 5 to 30 percent and output coverage ranging from 4 to 20 percent. We measured medium to high coverage for the more recent small-scale systems: reach coverage ranging from 43 to 85 percent, output coverage ranging from 40 to 85 percent.
% RESULTS
% REACH COVERAGE and
% OUTPUT COVERAGE
\begin{figure}
	\centering	
%	\resizebox{0.4\textwidth}{!}{%
%		\input{diagrams/coverage}	
%	}
	\includegraphics[width=0.5\textwidth]{images-paper/coverage.eps}
	\caption{
		Results for output candidates: \emph{Coverage loss} repesents reached output candidates that were not part of the output. \emph{Simple file} and \emph{function definition} represent context classifications of missed output candidates.
	}
	\label{coverage}
\end{figure}

@@

% CONTEXT DISTRIBUTION
% - FUNCTION CONTEXT
% - TEMPLATE CONTEXT
%\begin{figure}
%	\resizebox{0.4\textwidth}{!}{%
%		\input{diagrams/context_distribution}
%	}	
%	\caption{Context distribution of never reached output candidates for the PHP
%	corpus.}
%	\label{fig:contexts}
%\end{figure}

@@

% INCLUDE COVERAGE / SUCCESS RATE
\begin{figure}
%    \centering
%    \begin{subfigure}[t]{0.5 \textwidth}    	
%        \centering        
%        % Include Coverage Plot        %
%		\input{diagrams/include_coverage}		
%        \caption{Proportion of reached include expressions per project.}
%        \label{fig:include_coverage_results1}
%    \end{subfigure}
%	\hfill
%    \begin{subfigure}[t]{0.5\textwidth}    
%        \centering
%        % Success Rate Plot        
%        \input{diagrams/success_rate}        
%        \caption{Rate of successful resolutions of reached include expressions
%        per project.}
%        \label{fig:include_coverage_results2}
%    \end{subfigure}

\includegraphics[width=0.5\textwidth]{images-paper/includes.eps}

    \caption{Include coverage and accuracy.\label{fig:include_coverage_results}}
\end{figure}

@@

% Callback candidate description
\begin{figure}[t!]
	\centering
	\includegraphics[scale=0.55]{images-paper/callgraph_cases}
    \caption{Callgraph analysis: The bottom nodes represent functions containing missed output candidates, the top nodes depict entry points to them.}
    \label{fig:callgraph}
\end{figure}

@@

For failed function calls to functions containing not-reached output candidates we measured how many of those functions were only, or partially accessibly through indirect calling mechanisms (see Section \ref{sec:inaccessible}). As Figure \ref{fig:output_candidate_explanation} illustrates, around 80 percent (except for \textsf{TimeClock}) all missed output candidates of each system were only accessible through callback candidates, i.e., functions that are only accessible through indirect calling mechanisms.

% CALLBACK CANDIDATE ACCESSIBILITY ANALYSIS
\begin{figure}[p]
	\input{diagrams/output_candidates}
	\caption{Distribution dead function output candidates that are partially or
	completely explained due to callback candidates.}
	\label{fig:output_candidate_explanation}
\end{figure}

% CALLBACK CANDIDATE ACCESSIBILITY ANALYSIS
\begin{figure}[p]
	\input{diagrams/context_distribution}
	\caption{Distribution dead function output candidates that are partially or
	completely explained due to callback candidates.}
	\label{fig:coveragestuff}
\end{figure}

\section{Lessons Learned}

>>>>

\usepackage{xcolor}
%\usepackage[dvipsnames]{xcolor}

\definecolor{lightgrey}{HTML}{EEEEEE}

@@

<?php
function make_titles($arg, $n) {
  $is = shuffle(array(1,2,3));
  while ($n < $is[0]) {
    echo "<h1>" . $arg . "</h1><br />";
    $n = $n + 1;
  }
}

@@

$greeting = "Hello " . $_POST['name'] .  "!";
if (!isset($_POST['name'])) {
  echo "No name entered!";
} else {
  echo $greeting;
}
?>
\end{lstlisting}
		\caption{Snippet of HTML and PHP server-side web application code.
		\label{fig:web_application}}
    \end{subfigure}

@@

// #repeat n < shuffle(is)
<h1>Headline</h1><br />
// #endrepeat

@@

\end{lstlisting}
		\caption{
			Approximation client paege output of the server-side web application code in
			Figure \ref{fig:web_application}. Variability as well as repetition annotations
			are represented by preprocessor directives.
			\label{fig:output_example}
		}
    \end{subfigure}
    \vspace{2mm}
    \caption{
    	Illustration of static ouptut approximation using symbolic
    	execution
    	\label{fig:code_example}
    }
\end{figure}

Figure \ref{fig:code_example} shows two code listings. The snippet of server-side web
application code in Figure \ref{fig:web_application} contains both HTML and PHP. First, the
function \texttt{make\_titles} defined in lines 7 – 13 prints the string
\texttt{\$arg} passed as an argument between zero and an arbitrary number of
times. Second, the function is indirectly called using the built-in function
\texttt{call\_user\_func} in line 15 with the string \texttt{Headline} and
\texttt{4} as arguments.  Finally, the input from the form in lines 1 – 4 is
used: The variable \texttt{\$greeting} consists of three concatenated strings,
\texttt{Hello} the value of input field \texttt{name}, and an exclamation mark.
For this server-side code snippet there are two possible client page output
variants: If the input field \texttt{name} is empty at runtime, line 19 is
executed and \texttt{No name entered!} is printed, or, if a name was entered,
line 21 is executed and the greeting constructed in line 17 is printed.

%C. Introduction to symbolic execution of PHP code
The following symbolic execution semantics for PHP features  based on \cite{Nguyen:2014:BCG:2635868.2635928}
illustrate the necessary modifications made to handle the extent of ambiguity
that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable output as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.
First, loops in the previous execution semantics are only executed once instead
of an arbitrary number of times. Since a loop condition can contain symbolic
values, it might not be feasible to determine when a loop terminates. Hence,
any output constructed during the single loop iteration is highlighted
(\texttt{repeat \ldots  endrepeat}) to possibly be repeated an arbitrary number
of times. The output in line 7 is marked to be repeated an arbitrary number of times, as the number
of loop iterations cannot be determined statically. Second, a similar procedure
applies to recursive functions. Once called, a recursive function is  also only
executed to recursion depth one per call, and any subsequent invocation of that
function returns a symbolic value. This, again, is due to the difficulty to
determine the exact recursion depth similar to the number of loop iterations.
Third, since the symbolic execution engine applies a static analysis, any form
of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is substituted by a symbolic value. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternatives
(\texttt{if \ldots else \ldots endif}), as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

@@

All three tools though have only been applied to systems that are small and not
maintained any more. Despite the functional benefits these tools provide to
developers, their practicality for large and modern systems has not been
studied.

@@

In the following we describe in detail our implementation procedure, practical
problems we encountered and whether these are limiting scalability conceptually
or just require en gineering effort to be tamed.

\subsection{Towards Scalable Output Approximation?}
We re-implement\footnote{\url{https://github.com/smba/oak}} the existing
symbolic execution semantics of \cite{Nguyen:2014:BCG:2635868.2635928} with the
goal to support modern large-scale dynamic web applications, or at least to
better understand possible limitations. For reference, we choose to aim to
support  \textsf{WordPress}
\footnote{\url{https://github.com/WordPress/WordPress }}, a popular Content
Management System of reasonable size that is still maintained. Moreover, our
implementation uses parts of
Quercus\footnote{\url{http://quercus.caucho.com/}}, a pure Java implementation
of the PHP interpreter, which is tested against the \textsf{WordPress} test
suite. We choose a test-driven approach for our implementation with regression
tests and continuous integration. Our test cases are based on both custom
examples for different language features as well as \textsf{SchoolMate}, a 
small-scale system, and, at least in parts, \textsf{WordPress}, whose output we
manually validate and then automatically re-check for each test run.

@@

\subsection{Experiment Setup} 
For the evaluation we symbolically executed a selection of PHP systems. We
selected a corpus of twelve PHP systems with regard to system size and recency
as the corpus of the case study for \cite{Nguyen:2014:BCG:2635868.2635928} only
contained small-scale systems that are not maintained any more. The full list
of PHP systems is shown in Table \ref{corpus}. Our selection of PHP systems
includes

%
\begin{compactitem}
	\item four small-scale systems that selected from the previously mentioned case
	study in order to compare our results with the previous symbolic execution
	engine \cite{Nguyen:2014:BCG:2635868.2635928},

	\item three recent large-scale systems selected from the case study corpus of
	\cite{Hills:2013:ESP:2483760.2483786}, and

	\item four small-scale systems selected from a list of recent Content
	Management Systems \cite{codegeekz}.
\end{compactitem}

As for for the last two cases, we limited our selection since the parser we
used did not support all language features used. For the experimental evaluation we symbolically executed per system each script file or entry point (files with a \texttt{.php}, \texttt{.inc}, \texttt{.bit} or \texttt{.module} extension). All following measurements are cumulated per system for each entry point.

\begin{table*}[t]
\centering 
	\begin{tabular}{lrp{4cm}rrrr}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{Classification} & \textbf{SLOC} &
	\textbf{\#files} & \textbf{\#OCs} & \textbf{\#includes}
	\\
	\midrule
	AddressBook & 8.2.5.2 &  & 51,907 & 239  & 1009 & 186\\
	SchoolMate & 1.5.4 &  & 8,118 & 65  & 853 & 88\\
	TimeClock & 1.04 &  & 20,800 & 63  & 7920 & 306\\
	WebChess & 1.0.0 &  & 5,219 & 28  & 470 & 56\\
	\midrule
	Drupal & 7.5.0 & CMS & 52,464 & 125  & 3569 & 749\\
	phpBB & 3.1.9 &  & 327,371 & 1,398  & 3606 & 206\\
	phpMyAdmin & 4.6.3 &  & 303,582 & 871  & 7103 & 571\\
	\midrule
	Anchor & 0.12.1 &  & 15,054 & 201 & 987 & 32\\
	Kirby & XXX & CMS & XXX & XXX  & 654 & 23\\
	Automad & XXX & CMS & XXX & XXX  & 655 & 8\\
	Monstra & XXX & CMS & XXX & XXX  & 1934 & 48\\
	Nibbleblog & XXX & CMS & XXX & XXX  & 1013 & 28\\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

The coverage results are illustrated in Figure \ref{coverage}. We could
replicate high coverage results for the four small-scale systems with a reach
coverage and output coverage over 80 percent respectively. For the three
modern/large-scale systems we measured poor coverage with reach coverage
ranging from 5 to 30 percent and output coverage ranging from 4 to 20 percent.
We measured medium to high coverage for the more recent small-scale systems:
reach coverage ranging from 43 to 85 percent, output coverage ranging from 40
to 85 percent.

@@

@@

@@

% Callback candidate description
\begin{figure}[t!]
	%\centering
	\includegraphics[scale=0.55]{images-paper/callgraph_cases}
    \caption{Callgraph analysis: The bottom nodes represent functions containing missed output candidates, the top nodes depict entry points to them.}
    \label{fig:callgraph} output candidates we measured how many of those functions were only, or partially accessibly through indirect calling mechanisms (see Section \ref{sec:inaccessible}). As Figure \ref{fig:output_candidate_explanation} illustrates, around 80 percent (except for \textsf{TimeClock}) all missed output candidates of each system were only accessible through callback candidates, i.e., functions that are only accessible through indirect calling mechanisms.

\end{figure}

@@

For failed function calls to functions containing not-reached

	% RESULT DIAGRAMS
\begin{figure}[h!]
	\begin{subfigure}[center]{0.48\textwidth}
		\input{new_diagrams/not_reached}
		\caption{\label{coverage}}
	\end{subfigure}

	\begin{subfigure}[center]{0.48\textwidth}
		\input{new_diagrams/includes}
		\caption{
			Include coverage and accuracy.
			\label{fig:include_coverage_results}
		}
	\end{subfigure}

	% Callback candidate explanation
	\begin{subfigure}[center]{0.48\textwidth}
		\input{new_diagrams/callbacks}
		\caption{
			Distribution dead function output candidates that are par-
			tially or completely explained due to callback candidates.
			\label{fig:output_candidate_explanation}
		}

	\end{subfigure}
	\caption{Results}
\end{figure}
\section{Lessons Learned}
